A weighted sampling algorithm for the design of RNA sequences with targeted secondary structure and nucleotides distribution

### 2013-03-12 ISMB/results_ISMB.tex
more

\begin{figure*}[t!]
	\centering
 	\includegraphics[width=\textwidth]{Figures/time_rnastrand_clustered_rnainverse_100samples_fix}
	\caption{Average time in seconds to generate one sequence. We explicitly show 
	the time spent by \ourprog (full line) and \texttt{RNAinverse} (dashed) for various \GCContent. The first plot is in function
	of the length of the structures. The second is in function of the stack
	density (i.e. $2\cdot\#stacks/length$) and the latter in function of 
	the proportion of free nucleotides (i.e. unpaired bases) within loops.}
	\label{fig:time}	
\end{figure*}

@@

\begin{figure*}[ht!]	
	\centering
	\includegraphics[width=\textwidth]{Figures/mfe_struct_solve_nornainverse.png}
	\caption{Quality of \ourprog results. The first row shows the percentage
	of sampled sequences folding into the target. The second shows the 	
	proportion	of structures for which at least one correct sequence was 
	sampled.}
	\label{fig:mfe_struct_solved_noinverse}	
\end{figure*}

\begin{figure*}[ht!]	
	\centering
	\includegraphics[width=\textwidth]{Figures/mfe_struct_solved}
	\caption{Quality of \texttt{Incarnation+RNAinverse} results. The first row shows the percentage
	of sampled sequences folding into the target. The second shows the 	
	proportion	of structures for which at least one correct sequence was 
	sampled.}
	\label{fig:mfe_struct_solved}	
\end{figure*}

@@

A desirable feature in sequence design, is to produce samples with a high
diversity but with a conserved structure.
Two useful measures are the positional entropy and the base pair entropy.
The positional entropy is defined as:
\begin{align}
  \sum_{\substack{i\\b \in (\Ab,\Cb,\Gb,\Ub)}} -p_{i\to b}\cdot \log_4(p_{i\to b}) &&& \text{(Positional entropy)}
  \label{eq:entropy}
\end{align}
where $p_{i\to b}$ is the proportion of nucleotides $b$ at position $i$.
Intuitively, this measure captures the variability of sequences generated by a given method.
The base pair entropy is defined as:
\begin{align}
  \sum_{i<j}-p_{i,j}\cdot \log_2(p_{i,j})&&& \text{(Base-Pair entropy)}
  \label{eq:bpentropy}
\end{align}
where $p_{i,j}$ is the Boltzman probability, i.e. probability that a random structure drawn with respect to 
the Boltzmann distribution forms a base-pair between positions $i$ and $j$. The base pair entropy reflects
the distribution of structures weights in the Boltzmann ensemble. Thus, a low base pair entropy implies a highly conserved structure.
To compute the base pairs probabilities, we used \rnafold from the \texttt{Vienna Package 2.0} developed by~\cite{Hofacker:1994}.

  Figure~\ref{fig:nb_sols_entropy} shows the number of correct
solutions as their average entropy and base pair entropy, since 
\ourprog  only influences the distribution of nucleotides inside 
stacked base pairs. The most constraining factor for generating valid
 solutions seems to be  low \GCContent and the percentage of free nucleotides.

@@

\begin{figure*}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/nb_sols_entropy.png}
	\caption{Number of solutions generated and their average entropy. 
	The last row presents only the entropy of the nucleotides inside base 
	pairs.}
	\label{fig:nb_sols_entropy}
\end{figure*}

@@

It should be noted, however, than \RNAinverse produces sequences whose \GCContent is extremely concentrated (see Table~\ref{tab:nb_rnainv} in Supplementary Material), thereby requiring a large number of runs before producing a suitable \GCContent sequence. Therefore, a rejection algorithm based on \RNAinverse alone should not be considered a realistic solution for the design with constrained \GCContent.

>>>>

\begin{figure}[ht!]
	\centering
 	\includegraphics[width=0.5\textwidth]{Figures/time_incar_rnassd}
	\caption{
  Average time in seconds to generate one sequence for \ourprog and \RNAinverse.}
	%the time spent by \ourprog (full line) and \texttt{RNAinverse} (dashed) for various \GCContent. The first plot is in function
	%of the length of the structures. The second is in function of the stack
	%density (i.e. $2\cdot\#stacks/length$) and the latter in function of 
	%the proportion of free nucleotides (i.e. unpaired bases) within loops.}
	\label{fig:time}	
\end{figure}

@@

\begin{figure*}[ht!]	
	\centering
	\includegraphics[width=\textwidth]{Figures/mfe_struct_solved_vsrnainverse}
	\caption{Quality of \ourprog results. The first row shows the percentage
	of sampled sequences folding into the target when using only \ourprog. The second shows 
  after processing previous results with \RNAinverse.
	}
	\label{fig:mfe_struct_solved_noinverse}	
\end{figure*}

%\begin{figure*}[ht!]	
%	\centering
%	\includegraphics[width=\textwidth]{Figures/mfe_struct_solved}
%	\caption{Quality of \texttt{Incarnation+RNAinverse} results. The first row shows the percentage
%	of sampled sequences folding into the target. The second shows the 	
%	proportion	of structures for which at least one correct sequence was 
%	sampled.}
%	\label{fig:mfe_struct_solved}	
%\end{figure*}

@@

Using the same dataset of $50$ structures, we generated $100$ samples
per structure with classical softwares who can not control the \GC content.%with \RNAinverse. They yield for most parameters
%a high entropy.
%Table.~\ref{tab:nb_rnainv} contains the number of samples with a given \GCContent generated
with  \RNAinverse, \INFORNA, \NUPACK and \frankenstein.  We present in Fig.~\ref{fig:gcdist} those distributions. They have a highly concentrated distribution of \GC content showing one of theirs major deficiencies, with no sequence below $30\%$\GC. Therefore the need to be able to explicitly
 control the \GC content.

\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=0.5\textwidth]{Figures/histograme_5_gc_distribution_nornaexinv.png}
  \end{center}
  \caption{Overall \GCContent distribution for sequences designed using \RNAinverse, \INFORNA, \NUPACK and \frankenstein folding in the desired structure.}
  \label{fig:gcdist}
\end{figure}

A desirable feature in sequence design, is to produce samples with a high
diversity.
Two useful measures are the positional entropy and the sequence identity. %base pair entropy.
The positional entropy is defined as:
\begin{align}
  \sum_{\substack{i\\b \in (\Ab,\Cb,\Gb,\Ub)}} -p_{i\to b}\cdot \log_4(p_{i\to b}) &&& \text{(Positional entropy)}
  \label{eq:entropy}
\end{align}
where $p_{i\to b}$ is the proportion of nucleotides $b$ at position $i$.
Intuitively, this measure captures the variability of sequences generated by a given method.
The positional entropy is defined over a set $\mathcal{S}$ of sequences of equal length as follows:
\begin{align}
  \sum_{s^1,s^2 \in\mathcal{S}\times\mathcal{S}}\left(\frac{1}{|s^1|}{\sum_{\substack{i\\s^1_i\equiv s^2_i}}1}\right) &&& \text{Seq. identity}
\end{align}
where $s_i$ is the nucleotide at position $i$ in sequence $s$. The sequence identity is the ratio of perfectly conserved positions over
all pairwise sequences in the set. It reflects how over all the sequences the same positions are conserved. 
%The base pair entropy is defined as:
%\begin{align}
%  \sum_{i<j}-p_{i,j}\cdot \log_2(p_{i,j})&&& \text{(Base-Pair entropy)}
%  \label{eq:bpentropy}
%\end{align}
%where $p_{i,j}$ is the Boltzman probability, i.e. probability that a random structure drawn with respect to 
%the Boltzmann distribution forms a base-pair between positions $i$ and $j$. The base pair entropy reflects
%the distribution of structures weights in the Boltzmann ensemble. Thus, a low base pair entropy implies a highly conserved structure.
%To compute the base pairs probabilities, we used \rnafold from the \texttt{Vienna Package 2.0} developed by~\cite{Hofacker:1994}.

  Figure~\ref{fig:nb_sols_entropy} shows the number of correct
solutions as their average entropy and positions inside base pairs entropy, since 
\ourprog  only influences the distribution of nucleotides inside 
stacked base pairs. The most constraining factor for generating valid
 solutions seems to be  low \GCContent and the percentage of free nucleotides.

@@

\begin{figure*}[ht!]
  \centering
  \includegraphics[width=\textwidth]{Figures/seq_identity_10_30}
  \caption{Sequence identity of \ourprog and \RNAinverse for $10$ and $30\%$ of \GC.}
\end{figure*}

\begin{figure*}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/nb_sols_entropy}
	\caption{Number of solutions generated and their average entropy. 
	The last row presents only the entropy of the nucleotides inside base 
	pairs.}
	\label{fig:nb_sols_entropy}
\end{figure*}

@@

It should be noted, however, than \RNAinverse produces sequences whose \GCContent is extremely concentrated (see Fig.), thereby requiring a large number of runs before producing a suitable \GCContent sequence. Therefore, a rejection algorithm based on \RNAinverse alone should not be considered a realistic solution for the design with constrained \GCContent.

### 2013-03-12 ISMB/supplementary_ISMB.tex
more

%!TEX root = main_ISMB.tex
\section{Supplementary data}
\subsection{Benchmark \ourprog vs \RNAinverse}
Using the same dataset of $50$ structures, we generated $100$ samples
per structure %with \RNAinverse. They yield for most parameters
%a high entropy.
%Table.~\ref{tab:nb_rnainv} contains the number of samples with a given \GCContent generated
with  \RNAinverse, \INFORNA, \NUPACK and \frankenstein. They have a highly concentrated distribution of \GC content showing one of theirs major deficiencies. We present in Fig.~\ref{fig:gcdist} those distributions.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=0.5\textwidth]{Figures/histograme_5_gc_distribution_nornaexinv.png}
  \end{center}
  \caption{Overall \GCContent distribution for sequences designed using \RNAinverse, \INFORNA, \NUPACK and \frankenstein folding in the desired structure.}
  \label{fig:gcdist}
\end{figure}

 For all structures that have been solved 
by the three methods, only \RNAinverse, only \texttt{Incarnation} and
\texttt{Incarnation} followed by \RNAinverse, 
we present for every different concentration of \GCContent
the average sample entropy and the average base pairs entropy in Figure.~\ref{fig:rnainverse}.

%\begin{table}[h!]
%	\begin{center}
%		\begin{tabular}{|c|ccccc|}
%		\hline
%		Target \GCContent & 10 & 30 & 50 & 70 & 90\\ \hline
%   $\#$\RNAinverse samples& 11 & 437 & 3362 & 1155 & 35\\ \hline
%		\end{tabular}
%	\end{center}
%  \caption{Overall \GCContent distribution for sequences designed using \RNAinverse.}
%	\label{tab:nb_rnainv}
%\end{table}

\begin{figure*}[ht!]
	\centering
	%\hspace{-5em}
	\includegraphics[width=0.5\textwidth]{Figures/RNAinverse_data_100.png}
	\caption{Entropy and \GCContent  for structures solved by
	the 3 methods.}
	\label{fig:rnainverse}
\end{figure*}
{\em Récupérer Base-pair entropy}

%The same analysis but only for structures solved by \texttt{Incarnation}
%and \texttt{Incarnation} followed by \RNAinverse, is presented in 
%Fig.~\ref{fig:inc_rnainv}
%
%\begin{figure}
%	\centering
%	\includegraphics{}
%	\caption{Entropy and \texttt{C+G} content for structures solved by
%	the \texttt{Incarnation} and \texttt{Incarnation}$+$\RNAinverse.}
%	\label{fig:inc_rnainv}
%\end{figure}
\subsection{Limited impact on \GC of local-search postprocessing of \ourprog output}
Since local search approaches tend to experience a bias towards \GC{}-rich regions, it could be expected that our glocal approach, by postprocessing unpaired regions using a local search algorithm, would suffer from such a drift.
However, as summarized in Table~\ref{table:impact_on_gc}, we observed that the local search heuristic used to design nucleotides in loop regions has a very limited impact on the \GCContent. For each class of \GCContent, we reported the observed \GCContent in the sequence initially generated by \ourprog, and the observed \GCContent after the \RNAinverse postprocessing (as defined in Section \ref{subsec:glocal_method}). Our results show that the \GCContent is relatively well conserved (less than 6\% variation), with a general tendency of the postprocessing step to bring the \GCContent back to 50\%.

>>>>

%!TEX root = main_ISMB.tex
\section{Supplementary data}

\subsection{Benchmark \ourprog+\RNAinverse}
To emphasize the usefulness of processing \ourprog sequences with \RNAinverse, we present the number
of structures for which at least one sequence was generated with the desired MFE in Figure.~\ref{fig:nb_sols}

\begin{figure*}[ht!]
  \centering
  \includegraphics[width=\textwidth]{Figures/struct_solved_vsrnainverse.png}
  \caption{The first row shows the number of structures for which one generated sequence has the 
  structure as MFE when only using \ourprog. The second row shows when we process \ourprog results with \RNAinverse.}
  \label{fig:nb_sols}
\end{figure*}

\subsection{Benchmark \ourprog vs \RNASSD}
%Using the same dataset of $50$ structures, we generated $100$ samples
%per structure %with \RNAinverse. They yield for most parameters
%a high entropy.
%Table.~\ref{tab:nb_rnainv} contains the number of samples with a given \GCContent generated
%with  \RNAinverse, \INFORNA, \NUPACK and \frankenstein. They have a highly concentrated distribution of \GC content showing one of theirs major deficiencies. We present in Fig.~\ref{fig:gcdist} those distributions.

%\begin{figure}[ht]
%  \begin{center}
%    \includegraphics[width=0.5\textwidth]{Figures/histograme_5_gc_distribution_nornaexinv.png}
%  \end{center}
%  \caption{Overall \GCContent distribution for sequences designed using \RNAinverse, \INFORNA, \NUPACK and \frankenstein folding in the desired structure.}
%  \label{fig:gcdist}
%\end{figure}

 For all structures that have been solved 
by the three methods, only \RNASSD, only \texttt{Incarnation} and
\texttt{Incarnation} followed by \RNAinverse, 
we present for different concentration of \GCContent the average sequence identity in Figure.~\ref{fig:identity_50_70_90}
%the average sample entropy and the average base pairs entropy in Figure.~\ref{fig:rnainverse}.

\begin{figure*}[ht!]
  \centering
  \includegraphics[width=\textwidth]{Figures/seq_identity_50_70_90.png}
  \caption{Sequence identity by average \GCContent of $50$, $70$ and $90\%$ for \ourprog,\ourprog+\RNAinverse and \RNASSD.}
  \label{fig:identity_50_70_90}
\end{figure*}
%\begin{table}[h!]
%	\begin{center}
%		\begin{tabular}{|c|ccccc|}
%		\hline
%		Target \GCContent & 10 & 30 & 50 & 70 & 90\\ \hline
%   $\#$\RNAinverse samples& 11 & 437 & 3362 & 1155 & 35\\ \hline
%		\end{tabular}
%	\end{center}
%  \caption{Overall \GCContent distribution for sequences designed using \RNAinverse.}
%	\label{tab:nb_rnainv}
%\end{table}

%\begin{figure*}[ht!]
%	\centering
	%\hspace{-5em}
%	\includegraphics[width=0.5\textwidth]{Figures/RNAinverse_data_100.png}
%	\caption{Entropy and \GCContent  for structures solved by
%	the 3 methods.}
%	\label{fig:rnainverse}
%\end{figure*}
%{\em Récupérer Base-pair entropy}

%The same analysis but only for structures solved by \texttt{Incarnation}
%and \texttt{Incarnation} followed by \RNAinverse, is presented in 
%Fig.~\ref{fig:inc_rnainv}
%
%\begin{figure}
%	\centering
%	\includegraphics{}
%	\caption{Entropy and \texttt{C+G} content for structures solved by
%	the \texttt{Incarnation} and \texttt{Incarnation}$+$\RNAinverse.}
%	\label{fig:inc_rnainv}
%\end{figure}

\subsection{Limited impact on \GC of local-search postprocessing of \ourprog output}
Since local search approaches tend to experience a bias towards \GC{}-rich regions, it could be expected that our glocal approach, by postprocessing unpaired regions using a local search algorithm, would suffer from such a drift.
However, as summarized in Table~\ref{table:impact_on_gc}, we observed that the local search heuristic used to design nucleotides in loop regions has a very limited impact on the \GCContent. For each class of \GCContent, we reported the observed \GCContent in the sequence initially generated by \ourprog, and the observed \GCContent after the \RNAinverse postprocessing (as defined in Section \ref{subsec:glocal_method}). Our results show that the \GCContent is relatively well conserved (less than 6\% variation), with a general tendency of the postprocessing step to bring the \GCContent back to 50\%.
